### 2024-09-19 teaching/math240f24/draft.tex
Sophia Howard created

\medskip\boldlabel Proofs by contradiction. We now discuss a powerful method of proof, which dates
back to at least the
ancient Greeks and referred to for much of Western history by its Latin name,
{\it reductio ad absurdum}. It proceeds to prove
a statement $p$ by assuming its negation $\neg p$
holds, then deriving a contradiction, i.e., a statement whose truth value is $0$.
By doing this, one will have proved that $\neg p\To 0$, which in turn shows that $p$ is true, since
$$\neg p \To 0 \equiv \neg(\neg p) \vee 0 \equiv p\vee 0 \vee p.$$
Here is an example of a proof by contradiction.

@@

If $n$ is odd, then $n=2k+1$ for some integer $k$, and
$$1 + (-1)^{2k+1}\bigl(2(2k+1) - 1\bigr) = 1- (4k+2-1) = 1-(4k-1) = -4k,$$
which is also a multiple of $4$.\slug
\bye

Sometimes, the cases in which to split the proof are not so obvious.

@@

\medskip\boldlabel\datestamp{7}{} Mathematical induction. Sometimes we want to prove an infinite
number of statements, indexed by the natural numbers. If the complexity of the statements
sort of ``grows'' in $n$ (a vague notion to be made precise soon), the following theorem holds.

@@

Here are some examples and non-examples.
\medskip
\item{i)} The function $f:\RR\to\RR$ given by $f(x) = x^2$ is
$$f = \bigl\{ (a,b)\in \RR^2 : b = a^2 \bigl\},$$
when written in set-builder notation.
\smallskip
\item{ii)} On the other hand, the set
$$g = \bigl\{ (a,b)\in \RR^2 : a = b^2 \bigl\}$$
is not a function, since $(1,1)$ and $(1,-1)$ are both in $g$. Furthermore,
there is no element of $g$ with $-1$ as its first coordinate.
\smallskip
\item{iii)} If $X = \{1,4,5\}$ and $Y = \{0,1,2,3,4,5\}$, then
$$\bigl\{ (1,0), (2, 4), (5,5)\bigr\}$$
is a function.
\smallskip
\item{iv)} The set
$$h = \bigl\{ (x,y)\in \RR^2 : y = 1/x \bigl\}$$
is not a function, since there is no $y$ such that $(0,y)\in h$. However
if we amend the domain and consider
$$h = \bigl\{ (x,y)\in \bigl(\RR\setminus \{0\}\bigr)\times \RR : y = 1/x \bigl\},$$
then in fact, $h:\RR\setminus \{0\} \to \RR$ is a function.
\medskip

>>>>

\medskip\boldlabel Proofs by contradiction. We now discuss a powerful method of proof, which dates
back to at least the
ancient Greeks and referred to for much of Western history by its Latin name,
{\it reductio ad absurdum}. It proceeds to prove
a statement $p$ by assuming its negation $\neg p$
holds, then deriving a contradiction, i.e., a statement whose truth value is $0$.
By doing this, one will have proved that $\neg p\To 0$, which in turn shows that $p$ is true, since
$$\neg p \To 0 \equiv \neg(\neg p) \vee 0 \equiv p\vee 0 \equiv p.$$
Here is an example of a proof by contradiction.

@@

If $n$ is odd, then $n=2k+1$ for some integer $k$, and
$$1 + (-1)^{2k+1}\bigl(2(2k+1) - 1\bigr) = 1- (4k+2-1) = 1-(4k+1) = -4k,$$
which is also a multiple of $4$.\slug

\noindent
\datestamp{19}{ix} \hskip\parindent Sometimes, the cases into which one might split the proof are not so obvious.

@@

\medskip\boldlabel Mathematical induction. Sometimes we want to prove an infinite
number of statements, indexed by the natural numbers. If the complexity of the statements
sort of ``grows'' in $n$ (a vague notion to be made precise soon), the following theorem holds.

@@

Here are some examples and non-examples.
\medskip
\item{i)} The function $f:\RR\to\RR$ given by $f(x) = x^2$ is
$$f = \bigl\{ (a,b)\in \RR^2 : b = a^2 \bigl\},$$
when written in set-builder notation.
\smallskip
\item{ii)} On the other hand, the set
$$g = \bigl\{ (a,b)\in \RR^2 : a = b^2 \bigl\}$$
is not a function, since $(1,1)$ and $(1,-1)$ are both in $g$. Furthermore,
there is no element of $g$ with $-1$ as its first coordinate.
\smallskip
\item{iii)} If $X = \{1,2,5\}$ and $Y = \{0,1,2,3,4,5\}$, then
$$\bigl\{ (1,0), (2, 4), (5,5)\bigr\}$$
is a function from $X$ to $Y$.
\smallskip
\item{iv)} The set
$$h = \bigl\{ (x,y)\in \RR^2 : y = 1/x \bigl\}$$
is not a function, since there is no $y$ such that $(0,y)\in h$. However
if we amend the domain and consider
$$h = \bigl\{ (x,y)\in \bigl(\RR\setminus \{0\}\bigr)\times \RR : y = 1/x \bigl\},$$
then in fact, $h:\RR\setminus \{0\} \to \RR$ is a function.
\medskip
\bye
